---
title: "Univariate EDA in R"
author: "Caleb Lee"
format: html
execute:
    echo: true
---

## Load Packages & Data
```{r}
library(tidyverse) # tidyverse package (include ggplot2, dplyr, readr, etc.) - modern R tools for data import, wrangling, and plotting.

bigfive <- read_csv('./data/cleaned/cleaned_bigfive_v1.csv') # read_csv() - reads csv into an R data frame

#Sampling
set.seed(12) # same random seed for reproducibility
bigfive_sample <- bigfive %>%
  sample_n(5000)

glimpse(bigfive) # glimpse() - quick overview 
summary(bigfive) # summary() - basic descriptive stats
```
### Pivot to long format for easy faceting
```{r}
# Pivot to long format (makes faceting easy)
bigfive_long <- bigfive_sample %>%
  pivot_longer(
    cols = c(Openness, Conscientiousness, Extraversion, Agreeableness, Neuroticism),
    names_to = 'Trait',
    values_to = 'Score'
  )
```
## Descriptive Stats in knitr::kable
```{r}
# Descriptive Stats 
trait_summary <- bigfive_long %>%
  group_by(Trait) %>%
  summarize(
    n = n(),
    mean = mean(Score, na.rm = TRUE),
    sd = sd(Score, na.rm = TRUE), 
    min = min(Score, na.rm = TRUE), 
    max = max(Score, na.rm = TRUE)
  )
knitr::kable(trait_summary, caption = "Table 1. Descriptive Statistics for Big Five Traits (n = 5,000)")
```
## Trait Histograms
### Single trait(Extraversion) histogram + KDE + mean line 
```{r}
# Quick test: single trait histogram + KDE + mean line, fixed Likert 1–5

# Calculate mean
extraversion_mean <- mean(bigfive_sample$Extraversion, na.rm = TRUE)

# Histogram + KDE + mean line
ggplot(bigfive_sample, aes(x = Extraversion)) + 
  geom_histogram(
    aes(y = ..density..),
    bins = 30,
    fill = 'steelblue',
    color = 'white'
  ) +
  geom_density(
    color = 'red',
    size = 1
  ) +
  geom_vline(
    xintercept = extraversion_mean,
    color = 'black',
    linetype = 'dashed', 
    size = 1
  ) +
  scale_x_continuous(limits = c(1, 5), breaks = 1:5) +
  labs(
    title = "Histogram + KDE: Extraversion (n = 5,000)",
    x = 'Score',
    y = 'Density'
  ) +
  theme_minimal()
```
### All trait histogram + KDE + mean line
```{r}
# All traits: facet histogram + KDE, same Likert scale

# Calculate means for each trait
trait_means <- bigfive_long %>%
  group_by(Trait) %>%
  summarize (Mean = mean(Score, na.rm = TRUE))

# Faceted histogram + KDE + mean line 
ggplot(bigfive_long, aes(x = Score)) +
  geom_histogram(
    aes(y = ..density..),
    bins = 30,
    fill = 'steelblue',
    color = 'white'
  ) +
  geom_density(
    color = 'red',
    size = 1
  ) +
  geom_vline(
    data = trait_means,
    aes(xintercept = Mean),
    color = 'black',
    linetype = 'dashed',
    size = 1
  ) +
  facet_wrap(~ Trait) +
  scale_x_continuous(limits = c(1, 5), breaks = 1:5) +
  labs(
    title = "Big Five Traits: Histogram + KDE (n = 5,000)",
    x = 'Score',
    y = 'Density'
  ) +
  theme_minimal()
```

## Boxplot for Outlier Checks
### Single Boxplot (Openness)
```{r}
ggplot(bigfive_sample, aes(y = Openness, x = '', fill = 'lightblue')) +
  geom_boxplot(fill = 'lightblue') +
  geom_hline(
    yintercept = mean(bigfive_sample$Openness, na.rm = TRUE),
    color = 'red',
    linetype = 'dashed',
    size = 1
  ) +
  scale_y_continuous(limits = c(1,5), breaks = 1:5) +
  labs(
    title = "Boxplot: Openness with Mean Line (n = 5,000)",
    x = '', 
    y = 'Openness Score'
  ) +
  theme_minimal()
```
### Boxplot for all traits
```{r}
ggplot(bigfive_long, aes(x = Trait, y = Score)) + 
  geom_boxplot(fill = 'lightblue') + 
  stat_summary (fun = mean, geom = 'point', shape = 20, size = 3, color = 'red') +
  scale_y_continuous(limits = c(1,5), breaks = 1:5) + 
  labs (
    title = 'Boxplots of Big Five Traits with Mean Points',
    x = 'Trait',
    y = 'Score'
  ) +
  theme_minimal()