--- 
title: "Group Coparion in R without Sampling, using the full dataset"
author: "Caleb Lee"
format: html
execute: 
    echo: true
---
```{r}
# Import libraries
library(tidyverse)
library(broom)
bigfive <- read_csv('./data/cleaned/cleaned_bigfive_v1.csv')

# filtering for NA and NONE
bigfive_filtered <- bigfive %>%
    filter(!is.na(country) & country != 'NONE')

# seeing which countries are present
bigfive_filtered %>%
    count(country, sort = TRUE) %>%
    print (n = Inf)
```

```{r}
# created a CSV file summarizing the name, row count, geographical and cultural location of all 220 countries present (after filtered for NA and NONE)
country_info <- read_csv('./data/cleaned/country_names_220.csv')

# show unique continents
country_info %>%
  count(continent, sort = TRUE)

# Show unique cultural groups
country_info %>%
  count(cultural_group, sort = TRUE)

# Sum total row_count by CONTINENT
country_info %>%
  group_by(continent) %>%
  summarize(
    n_countries = n(),
    total_row_count = sum(row_count, na.rm = TRUE)
  ) %>%
  arrange(desc(total_row_count))

# Sum total row_count by CULTURAL GROUP
country_info %>%
  group_by(cultural_group) %>%
  summarize(
    n_countries = n(),
    total_row_count = sum(row_count, na.rm = TRUE)
  ) %>%
  arrange(desc(total_row_count))

```
```{r}
# Grouping by continent:
# put russia in europe, put turkey in asia, and caribbean countries into south america
library(dplyr)

bigfive_continent <- bigfive_filtered %>%
  mutate(continent = case_when(
    # North America (mainland + Central America only)
    country %in% c("US", "CA", "MX", "PR", "HN", "NI", "CR", "PA", "GT", "SV") ~ "North America",
    
    # South America (+ Caribbean islands grouped here)
    country %in% c(
      "BR", "AR", "CO", "CL", "PE", "VE", "UY", "DO", "EC", "GY", "SR", "BO", "PY",
      "BZ", "JM", "TT", "GD", "AG", "BS", "CU", "DM", "KN", "LC", "VC", "HT", "BB", "CW", "KY", "AW", "SX", "MQ", "GP", "BM"
    ) ~ "South America",
    
    # Europe (+ Russia)
    country %in% c(
      "GB", "IE", "IS", "NO", "SE", "FI", "DK", "NL", "FR", "DE", "IT", "ES", "PL", "PT", "BE", "AT", "CH", "LU", "MT",
      "RU", "UA", "RS", "RO", "BG", "GR", "GE", "CY", "MK", "ME", "MD", "AM", "BY",
      "CZ", "HU", "SK", "SI", "HR", "AL", "BA", "LT", "LV", "EE",
      "LI", "MC", "SM", "GG", "JE", "IM", "FO", "AX", "XK", "AD"
    ) ~ "Europe",
    
    # Asia (+ Turkey)
    country %in% c(
      "IN", "PK", "BD", "LK", "NP", "MV", "CN", "HK", "JP", "KR", "TW", "SG", "MY", "PH", "VN", "TH", "KH", "ID", "BN", "LA",
      "MM", "TL", "PG", "MN", "KZ", "KG", "UZ", "AZ", "IL", "IQ", "SY", "IR", "JO", "LB", "PS", "YE", "SA", "AE", "KW", "QA",
      "BH", "OM", "MO", "TR"
    ) ~ "Asia",
    
    # Africa
    country %in% c(
      "ZA", "KE", "NG", "GH", "ZM", "UG", "SD", "ZW", "ET", "DZ", "MA", "TZ", "RW", "CM", "AO", "MW", "MZ", "SN", "SZ",
      "LS", "BI", "BF", "BJ", "GM", "GN", "CI", "CV", "GQ", "KM", "MG", "MR", "NE", "SL", "TG", "TD", "CD", "DJ", "CG", "GA",
      "SS", "EG", "LY"
    ) ~ "Africa",
    
    # Oceania
    country %in% c("AU", "NZ", "FJ", "WS", "VU", "CK", "FM", "PW", "PG", "TO", "TL", "MP", "GU", "NC", "MH", "NU") ~ "Oceania",
    
    # Other / Unclassified
    TRUE ~ "Other"
  ))


  bigfive_continent %>%
  count(continent) %>%
  arrange(desc(n))

