---
title: "Group Comparison in R"
author: "Caleb Lee"
format: html
execute:
    echo: true
---
## Creating continent groups
### Import libraries and sample
```{r}
library(tidyverse)
library (broom)
bigfive <- read_csv('./data/cleaned/cleaned_bigfive_v1.csv') # read_csv() - reads csv into an R data frame

#Sampling
set.seed(42) # same random seed for reproducibility
bigfive_sample <- bigfive %>%
  sample_n(5000)
```
### Creating Continent groups; filtering for NONE and minor countries.
```{r}
#Checking which countries are present
bigfive_sample %>%
    count(country, sort = TRUE) %>%
    print (n = Inf)
```

```{r}
# Sort countries into continents (same as python)

# Continent group creation
north_america <- c('US', 'CA', 'MX', 'PR', 'CR', 'DO', 'JM', 'TT', 'BM', 'HN')
south_america <- c('BR', 'AR', 'CL', 'CO', 'PE', 'VE', 'EC', 'BO', 'PY')
europe <- c( 'RO','GB', 'IE', 'DE', 'FR', 'NL', 'SE', 'NO', 'FI', 'DK', 'BE', 'PL',
             'IT', 'PT', 'ES', 'CH', 'AT', 'CZ', 'HU', 'SK', 'EE', 'LT', 'UA',
             'BG', 'GR', 'RS', 'HR', 'CY', 'MK', 'BA', 'IS', 'JE', 'LV', 'MD',
             'ME', 'AD', 'AL')
asia <- c('IN', 'PH', 'MY', 'SG', 'ID', 'PK', 'HK', 'JP', 'KR', 'VN', 'TH',
          'IR', 'LB', 'TW', 'AE', 'IL', 'JO', 'NP', 'CN', 'KH', 'QA', 'BN',
          'KZ', 'MO', 'OM', 'BD', 'IQ', 'GE', 'BH', 'KW', 'LK', 'SA')
oceania <- c('AU', 'NZ', 'PG', 'GU')

bigfive_sample <- bigfive_sample %>%
    # filtering NONE entries
    filter(country != 'NONE') %>%
    mutate(
        continent = case_when(
            country %in% north_america ~ 'NA',
            country %in% south_america ~ 'SA',
            country %in% europe ~ 'EU',
            country %in% asia ~ 'AS',
            country %in% oceania ~ 'OC',
            TRUE ~ 'Other'
        )
    ) 
bigfive_sample %>%
    count(continent, sort = TRUE)

```
```{r}
# What countries are in Other
bigfive_sample %>%
  filter(continent == 'Other') %>%
  count(country, sort = TRUE)
```
```{r}
# Excluded a bunch of African countries and transcontinental countries (RU, TR) due to low n
bigfive_sample <- bigfive_sample %>%
    filter(continent != 'Other')
```
### One-way ANOVA
```{r}
# Group means for Openness
bigfive_sample %>%
    group_by(continent) %>%
    summarize(
        mean = mean (Openness, na.rm = TRUE)
    )
# One-way ANOVA for Openness
openness_anova <- aov(Openness ~ continent, data = bigfive_sample)
anova_tidy <- tidy(openness_anova)
summary(openness_anova)
knitr::kable(anova_tidy, caption = "Table: ANOVA for Openness by Continent")

# Tukey HSD for Openness
TukeyHSD(openness_anova)
```

#### Review again tomorrow
```{r}
# One-way ANOVA result in knitr::kable:
# Fit ANOVA model
openness_anova <- aov(Openness ~ continent, data = bigfive_sample)

# Tidy ANOVA result
anova_tidy <- tidy(openness_anova)

# Display in table
knitr::kable(
  anova_tidy,
  caption = "Table 1. One-way ANOVA for Openness by Continent",
  digits = 4
)

# Tukey HSD in knitr::kable:
openness_tukey <- TukeyHSD(openness_anova)

# Convert to dataframe
tukey_df <- as.data.frame(openness_tukey$continent)

# Add comparison pair as a column
tukey_df$comparison <- rownames(tukey_df)

# Reorder columns for clarity
tukey_df <- tukey_df %>%
  select(comparison, diff, lwr, upr, `p adj`)

# Display nicely
knitr::kable(
  tukey_df,
  caption = "Table 2. Tukey HSD Post-Hoc Test for Openness by Continent",
  digits = 4
)
```
```{r}
# Looping for all traits
traits <- c('Openness', 'Conscientiousness', 'Extraversion', 'Agreeableness', 'Neuroticism')

for (trait in traits) {
    cat('\n\n### ANOVA for', trait, 'by Continent ###\n\n')

    # Group means
    means <- bigfive_sample %>%
        group_by(continent) %>%
        summarize(
            mean = mean(.data[[trait]], na.rm = TRUE)
        )
    print(means)

    # ANOVA
    model <- aov(as.formula(paste(trait, '~ continent')), data = bigfive_sample)
    print (summary(model))

    # Tukey HSD
    cat('\nTukey HSD:\n')
    print(TukeyHSD(model))
}
```
#### Review again tomorrow
```{r}
traits <- c('Openness', 'Conscientiousness', 'Extraversion', 'Agreeableness', 'Neuroticism')

for (trait in traits) {
  cat('\n\n### ANOVA for', trait, 'by Continent\n\n')

  # Group means
  means <- bigfive_sample %>%
    group_by(continent) %>%
    summarize(mean = mean(.data[[trait]], na.rm = TRUE))
  print(means)

  # ANOVA
  model <- aov(as.formula(paste(trait, '~ continent')), data = bigfive_sample)
  tidy_anova <- tidy(model)
  print(
    knitr::kable(
      tidy_anova,
      caption = paste('Table: ANOVA for', trait),
      digits = 4
    )
  )

  # Tukey HSD
  tukey <- TukeyHSD(model)
  tukey_df <- as.data.frame(tukey$continent)
  tukey_df$comparison <- rownames(tukey_df)
  tukey_df <- tukey_df %>%
    select(comparison, diff, lwr, upr, `p adj`)
  print(
    knitr::kable(
      tukey_df,
      caption = paste('Table: Tukey HSD for', trait),
      digits = 4
    )
  )
}
```
### ANOVA Result (Review later)
#### Openness:
ANOVA: F = 11.75, p < .001 → significant overall difference.

Tukey: Asia (AS) significantly lower than all other continents (EU, NA, OC, SA).

EU–AS, NA–AS, OC–AS, SA–AS: all p < .01.

Other pairs not significant.

Participants in Asia scored significantly lower on Openness than those from Europe, North America, Oceania, and South America. No other differences were significant.

#### Conscientiousness
ANOVA: F = 5.27, p < .001 → significant.

Tukey: Only SA–AS is significant (p = .005).

South America mean = 3.17 → Asia mean = 3.14 → SA slightly higher.

NA–AS, EU–AS, other pairs → NS.

South America showed slightly higher Conscientiousness than Asia. No other pairs were significant.

#### Extraversion
ANOVA: F = 8.13, p < .001 → significant.

Tukey:

EU–AS, NA–AS, SA–AS significant.

Other pairs → NS.

Asia scored significantly lower in Extraversion than Europe, North America, and South America.

####  Agreeableness
ANOVA: F = 14.93, p < .001 → significant.

Tukey:

EU–AS, NA–AS, OC–AS, SA–AS → all significant.

Other pairs → NS.

Asia scored significantly lower in Agreeableness than Europe, North America, Oceania, and South America.

####  Neuroticism
ANOVA: F = 11.69, p < .001 → significant.

Tukey:

NA–AS, OC–AS, SA–AS → significant.

SA–EU, SA–OC → significant.

Asia showed significantly lower Neuroticism than North America, Oceania, and South America. South America also scored significantly lower than Europe and Oceania. (Many pairs have small diffs)