--- 
title: "Correlation Matrix Analysis"
author: "Caleb Lee"
format: html
execute: 
    echo: true
---
## Objective
To explore the interrelationships among Big Five personality traits by computing and visualizing a correlation matrix based on reverse-scored trait scores.

## 1. Load Libraries and Data
```{r}
library(tidyverse)
library(psych)
library(ggcorrplot)

bigfive <- read_csv('./data/cleaned/cleaned_bigfive_v1.csv')
```
## 2. Identify Reverse Coded Items
```{r}
# running alpha() again to get reverse-scored items
get_reversed_items <- function(alpha_result) {
  keys <- unlist(alpha_result$keys)
  reversed_items <- keys[startsWith(keys, "-")]
  reversed_items <- gsub("-", "", reversed_items)
  return(reversed_items)
}

opn_items <- bigfive %>% select(matches("^OPN[0-9]+$"))
ext_items <- bigfive %>% select(matches("^EXT[0-9]+$"))
agr_items <- bigfive %>% select(matches("^AGR[0-9]+$"))
csn_items <- bigfive %>% select(matches("^CSN[0-9]+$"))
nrt_items <- bigfive %>% select(matches("^NRT[0-9]+$"))


alpha_opn <- alpha(opn_items, check.keys = TRUE)
alpha_ext <- alpha(ext_items, check.keys = TRUE)
alpha_agr <- alpha(agr_items, check.keys = TRUE)
alpha_csn <- alpha(csn_items, check.keys = TRUE)
alpha_nrt <- alpha(nrt_items, check.keys = TRUE)

# reversed-item names
reversed_opn <- get_reversed_items(alpha_opn)
reversed_ext <- get_reversed_items(alpha_ext)
reversed_agr <- get_reversed_items(alpha_agr)
reversed_csn <- get_reversed_items(alpha_csn)
reversed_nrt <- get_reversed_items(alpha_nrt)
```
```{r}
# checking
print(reversed_opn)
```
## 3. Applying Reverse Scoring
```{r}
# function to reverse code (-6) items
reverse_score <- function(df, items_to_reverse) {
    df %>% mutate(across(all_of(items_to_reverse), ~ 6 - .))
}

bigfive_scored <- bigfive %>%
    reverse_score(reversed_opn) %>%
    reverse_score(reversed_csn) %>%
    reverse_score(reversed_ext) %>%
    reverse_score(reversed_agr) %>%
    reverse_score(reversed_nrt)

```
## 4. Calculate Trait Scores
```{r}
bigfive_scored <- bigfive_scored %>%
    mutate(
        Openness = rowMeans(select(., matches("^OPN[0-9]+$"))),
        Conscientiousness = rowMeans(select(., matches("^CSN[0-9]+$"))),
        Extraversion = rowMeans(select(., matches("^EXT[0-9]+$"))),
        Agreeableness = rowMeans(select(., matches("^AGR[0-9]+$"))),
        Neuroticism = rowMeans(select(., matches("^NRT[0-9]+$")))
    )
```
## 5. Correlation Matrix (ggcorrplot)
```{r}
traits <- bigfive_scored %>%
    select(Openness, Conscientiousness, Extraversion, Agreeableness, Neuroticism)

cor_matrix <- cor(traits, use = 'pairwise.complete.obs', method = 'pearson')
print(round(cor_matrix, 2))
```
## 6. Visualization
```{r}
ggcorrplot(cor_matrix, method = 'circle', type = 'upper', lab = TRUE, lab_size = 4, colors = c('blue', 'white', 'red'), title = 'Correlation Matrix: Big Five Traits', ggtheme = theme_minimal())
```
## 7. Test for Significance
```{r}
cor_test <- corr.test(traits, use = 'pairwise.complete.obs', method = 'pearson', adjust = 'holm')

cor_test$r
cor_test$p
cor_test$p.adj

round(cor_test$r, 2)       
round(cor_test$p, 4)      
round(cor_test$p.adj, 4) 

# They are all strongly significant due to very large sample size (+602k)
```
### Correlation Matrix (corrplot)
```{r}
# Create a significance matrix (TRUE = p < .05)
library(corrplot)

corrplot(cor_test$r, 
         method = "color", 
         type = "upper", 
         p.mat = cor_test$p, 
         sig.level = 0.05, 
         insig = "blank", 
         addCoef.col = "black",
         tl.col = "black", 
         tl.srt = 45,
         title = "Correlation Matrix: Big Five Traits")

```
## 8. Interpretation + Inline Stats
```{r}
# Extract rounded values for inline reporting
n_used <- bigfive_scored %>%
  select(Openness, Conscientiousness, Extraversion, Agreeableness, Neuroticism) %>%
  drop_na() %>%
  nrow()

r_E_A <- round(cor_test$r["Extraversion", "Agreeableness"], 2)
p_E_A <- "< .001"

r_N_C <- round(cor_test$r["Neuroticism", "Conscientiousness"], 2)
r_N_E <- round(cor_test$r["Neuroticism", "Extraversion"], 2)
r_O_C <- round(cor_test$r["Openness", "Conscientiousness"], 2)
r_O_E <- round(cor_test$r["Openness", "Extraversion"], 2)
r_O_A <- round(cor_test$r["Openness", "Agreeableness"], 2)
r_O_N <- round(cor_test$r["Openness", "Neuroticism"], 2)
```

A Pearson correlation matrix was computed among the Big Five personality traits. All correlations were statistically significant (*p* < .001) due to the large sample size (*N* = `r n_used`). Extraversion showed a moderate positive correlation with Agreeableness (*r* = `r r_E_A`, *p* `r p_E_A`). Neuroticism correlated positively with both Conscientiousness (*r* = `r r_N_C`, *p* < .001) and Extraversion (*r* = `r r_N_E`, *p* < .001). Other correlations were weak although significant, with Openness showing small positive associations with Conscientiousness (*r* = `r r_O_C`), Extraversion (*r* = `r r_O_E`), Agreeableness (*r* = `r r_O_A`), and Neuroticism (*r* = `r r_O_N`).
