--- 
title: "Factor Analysis"
author: "Caleb Lee"
format: html
execute: 
    echo: true
---
## Objective
Uncover the underlying factor structure of the Big Five item-level data using Exploratory Factor Analysis (EFA).

## 1. Setup - Load libraries and data
```{r}
library(tidyverse)
library(psych)

bigfive <- read_csv('./data/cleaned/cleaned_bigfive_v1.csv')
```
## 2. Identify reverse-coded items
```{r}
# Function to extract reversed items
get_reversed_items <- function(alpha_result) {
  keys <- unlist(alpha_result$keys)
  reversed_items <- keys[startsWith(keys, "-")]
  reversed_items <- gsub("-", "", reversed_items)
  return(reversed_items)
}

# Subset each trait's item columns
opn_items <- bigfive %>% select(matches("^OPN[0-9]+$"))
ext_items <- bigfive %>% select(matches("^EXT[0-9]+$"))
agr_items <- bigfive %>% select(matches("^AGR[0-9]+$"))
csn_items <- bigfive %>% select(matches("^CSN[0-9]+$"))
nrt_items <- bigfive %>% select(matches("^NRT[0-9]+$"))

# Get reversed items using psych::alpha() and extract keys
alpha_opn <- alpha(opn_items, check.keys = TRUE)
alpha_ext <- alpha(ext_items, check.keys = TRUE)
alpha_agr <- alpha(agr_items, check.keys = TRUE)
alpha_csn <- alpha(csn_items, check.keys = TRUE)
alpha_nrt <- alpha(nrt_items, check.keys = TRUE)

reversed_opn <- get_reversed_items(alpha_opn)
reversed_ext <- get_reversed_items(alpha_ext)
reversed_agr <- get_reversed_items(alpha_agr)
reversed_csn <- get_reversed_items(alpha_csn)
reversed_nrt <- get_reversed_items(alpha_nrt)
```
## 3. Apply Reverse Scoring
```{r}
# Define reverse-scoring function
reverse_score <- function(df, items_to_reverse) {
  df %>% mutate(across(all_of(items_to_reverse), ~ 6 - .))
}

# Apply reverse scoring to create final scored dataset
bigfive_scored <- bigfive %>%
  reverse_score(reversed_opn) %>%
  reverse_score(reversed_ext) %>%
  reverse_score(reversed_agr) %>%
  reverse_score(reversed_csn) %>%
  reverse_score(reversed_nrt)
```
## 4. Subset item-level matrix for EFA
```{r}
item_data <- bigfive_scored %>%
  select(matches("^EXT[0-9]+$"),
         matches("^AGR[0-9]+$"),
         matches("^CSN[0-9]+$"),
         matches("^NRT[0-9]+$"),
         matches("^OPN[0-9]+$"))
```
## 5. Suitability for Factor Analysis
```{r}
# KMO - measure of sampling adequacy
KMO(item_data)
# Bartlett's test of sphericity
cortest.bartlett(cor(item_data), n = nrow(item_data))
```
## 6. Determine # of Factors
```{r}
# Scree plot and parallel analysis
fa.parallel(item_data, fa = 'fa', fm = 'ml')
```
## 7. Factor Analysis
```{r}
# Run EFA with 5 factors and oblimin rotation (allow correlated traits)
efa_result <- fa(item_data, nfactors = 5, rotate = "oblimin", fm = "ml")

# Print loadings (cutoff = 0.3)
print(efa_result$loadings, cutoff = 0.3)
```
```{r}
# Run EFA with 10 factors as suggested by parallel analysis
efa_result_10 <- fa(item_data, nfactors = 10, rotate = "oblimin", fm = "ml")
print(efa_result_10$loadings, cutoff = 0.3)

```
## 8. Visualization 
```{r}
# Factor diagram
fa.diagram(efa_result)
# Table of strong loadings
efa_loadings <- as.data.frame(efa_result$loadings[1:50, ])
efa_loadings %>% 
  filter(apply(abs(.), 1, max) >= 0.3) %>%
  round(2)
```
## 9. Results
This section reports the results of an exploratory factor analysis (EFA) conducted to investigate the latent structure of the 50-item Big Five personality inventory.

### 9.1 Data Preparation
Prior to EFA, reverse-coded items were identified using `alpha(..., check.keys = TRUE)` and recoded by subtracting item responses from 6 (i.e., `6 - x`). The final dataset (`item_data`) included all 50 Big Five items, grouped by trait: Openness (OPN1–10), Conscientiousness (CSN1–10), Extraversion (EXT1–10), Agreeableness (AGR1–10), and Neuroticism (NRT1–10).

### 9.2 Factorability Checks
```{r}
overall_kmo <- round(KMO(item_data)$MSA, 2)
bartlett <- cortest.bartlett(cor(item_data), n = nrow(item_data))
bartlett_chi <- round(bartlett$chisq)
bartlett_df <- bartlett$df
```
The Kaiser-Meyer-Olkin (KMO) measure of sampling adequacy was r overall_kmo, indicating excellent shared variance among items.
Bartlett’s test of sphericity was highly significant, χ²(`r bartlett_df`) = `r bartlett_chi`, p < .001, confirming that the item correlations were sufficient for factor analysis.

### 9.3 Number of Factors
A parallel analysis was conducted to determine the optimal number of factors to extract. The scree plot and simulated eigenvalues indicated support for a 10-factor solution, though a theory-driven 5-factor solution was also evaluated for comparison.
```{r}
cumulative_var_5 <- round(efa_result$Vaccounted["Cumulative Var", 5], 3) * 100
prop_var_5 <- round(efa_result$Vaccounted["Proportion Var", 1:5], 3)
cumulative_var_10 <- round(efa_result_10$Vaccounted["Cumulative Var", 10], 3) * 100
```

The 5-factor model accounted for `r cumulative_var_5`% of total variance, with individual contributions as follows:

Factor 1: `r prop_var_5[1]`

Factor 2: `r prop_var_5[2]`

Factor 3: `r prop_var_5[3]`

Factor 4: `r prop_var_5[4]`

Factor 5: `r prop_var_5[5]`

The 10-factor model explained slightly more (`r cumulative_var_10`%) variance and offered a more detailed view of subtraits.

### 9.4 Five-Factor Solution
```{r}
# Helper: compute the loading range (min/max abs loading) per trait
get_loading_range <- function(prefix, loadings) {
  items <- rownames(loadings)[grepl(paste0("^", prefix), rownames(loadings))]
  item_loadings <- loadings[items, ]
  max_loadings <- apply(abs(item_loadings), 1, max)
  round(range(max_loadings), 2)
}

# Calculate loading ranges for each trait (5-factor solution)
loading_ranges <- list(
  EXT = get_loading_range("EXT", efa_result$loadings),
  AGR = get_loading_range("AGR", efa_result$loadings),
  CSN = get_loading_range("CSN", efa_result$loadings),
  NRT = get_loading_range("NRT", efa_result$loadings),
  OPN = get_loading_range("OPN", efa_result$loadings)
)

# Proportion variance explained in 10-factor model
prop_var_10 <- round(efa_result_10$Vaccounted["Proportion Var", 1:10], 3)

# Extract top Openness loadings in 10-factor model
opn_items_10 <- rownames(efa_result_10$loadings)[grepl("^OPN", rownames(efa_result_10$loadings))]
opn_loadings_10 <- efa_result_10$loadings[opn_items_10, ]
top_opn_10 <- apply(abs(opn_loadings_10), 1, max)
top_opn_10 <- round(top_opn_10[c("OPN1", "OPN2", "OPN3", "OPN6", "OPN8")], 2)


```
The five-factor solution aligned closely with the canonical Big Five structure:

Factor 1: Extraversion (EXT1–EXT10)

Factor 2: Neuroticism (NRT1–NRT10)

Factor 3: Agreeableness (AGR1–AGR10)

Factor 4: Openness (OPN1–OPN10)

Factor 5: Conscientiousness (CSN1–CSN10)

Extraversion items showed factor loadings ranging from `r loading_ranges$EXT[1]` to `r loading_ranges$EXT[2]`, Neuroticism from `r loading_ranges$NRT[1]` to `r loading_ranges$NRT[2]`, Agreeableness from `r loading_ranges$AGR[1]` to `r loading_ranges$AGR[2]`, Conscientiousness from `r loading_ranges$CSN[1]` to `r loading_ranges$CSN[2]`, and Openness from `r loading_ranges$OPN[1]` to `r loading_ranges$OPN[2]`.

All 50 items loaded meaningfully (≥ .30) onto their expected factors, with minimal cross-loadings. This provides strong support for the structural validity of the Big Five model in this dataset.

### 9.5 Ten-Factor Solution (Facet-Level Insights)
A parallel analysis suggested a more complex 10-factor solution, which explained a slightly higher amount of variance (`r cumulative_var_10`%) than the five-factor model. The top five factors in this solution accounted for `r prop_var_10[1]`, `r prop_var_10[2]`, `r prop_var_10[3]`, `r prop_var_10[4]`, and `r prop_var_10[5]` of the total variance, respectively.

This solution revealed facet-level distinctions within broader traits:

Neuroticism appeared to divide into general emotional instability and a secondary anxiety-related cluster.

Openness separated into two subdimensions:

OPN3 (`r top_opn_10["OPN3"]`) and OPN6 (`r top_opn_10["OPN6"]`) loaded most strongly on an intellect-related factor.

OPN1 (`r top_opn_10["OPN1"]`) and OPN8 (`r top_opn_10["OPN8"]`) loaded highly on an aesthetic/imaginative factor.

OPN2 also showed a notable loading (`r top_opn_10["OPN2"]`), supporting this multidimensional structure.

These results suggest that the Big Five traits in this dataset may be composed of meaningful facet-level substructures, consistent with models like the NEO-PI-R. While more complex, this solution may offer deeper psychological insight and can inform future confirmatory work.



### 9.6 Factor Loading and Diagram
```{r}
efa_table <- as.data.frame(efa_result$loadings[1:50, ]) %>%
  filter(apply(abs(.), 1, max) >= 0.3) %>%
  round(2)

knitr::kable(efa_table, caption = "Five-Factor Loadings (≥ .30)")
```
```{r}
fa.diagram(efa_result)
```

A factor loading table (above) shows items with loadings ≥ .30. The diagram visually represents the structure of the 5-factor solution, confirming the clustering of items into their expected Big Five domains.


